---
title: "UCEC Molecular Subtype Classification: Final Reproducibility Report"
author: "Student ID: 3173204"
date: "31 July 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    self_contained: true
    df_print: paged
---

# Project purpose and rationale

Uterine corpus endometrial carcinoma is molecularly heterogeneous, and tumours
assigned to different genomic subtypes can differ in biological behaviour,
prognosis and potential treatment relevance. The central purpose of this
project was therefore to determine whether tumour transcriptomic profiles can
reliably distinguish three harmonised UCEC molecular groups—CN-high, CN-low
and Hypermutated—and to identify reproducible candidate genes associated with
those distinctions.

The analytical strategy combined complementary approaches because no single
method answers the full research question. Differential-expression analysis
was used to identify subtype-associated transcriptional changes; GO and KEGG
enrichment were used to investigate the biological processes represented by
those changes; and multinomial LASSO and Random Forest were used to test
whether expression patterns could predict molecular subtype in unseen samples.
Candidate biomarkers were prioritised by identifying genes supported by both
machine-learning approaches and then integrating differential-expression and
biological evidence.

TCGA-UCEC was used for model development and held-out internal evaluation,
while CPTAC-UCEC provided an independent external transfer assessment. This
design allowed the project to examine both internal predictive performance and
the extent to which the learned signal transferred across cohorts. The outputs
should be interpreted as computational evidence for candidate biomarkers and
subtype classification, rather than as clinical validation.

For reproducibility, the heavy workflow is executed through
`UCEC_Master_Analysis_Low_Memory.Rmd`. This report runs the later lightweight
figure and table scripts, displays the complete results, and provides
collapsible source code for every analytical stage.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)

options(stringsAsFactors = FALSE)

show_figure <- function(
    path,
    caption,
    width = "95%"
) {
  cat("\n\n")
  cat(
    paste0(
      "**",
      caption,
      "**\n\n"
    )
  )

  if (file.exists(path)) {
    cat(
      paste0(
        "![",
        caption,
        "](",
        path,
        "){width=",
        width,
        "}\n\n"
      )
    )
  } else {
    cat(
      paste0(
        "> Figure not found: `",
        path,
        "`. Run the associated script and knit again.\n\n"
      )
    )
  }
}

show_table <- function(
    path,
    caption,
    n = Inf,
    digits = 4
) {
  if (file.exists(path)) {
    table_data <- readr::read_csv(
      path,
      show_col_types = FALSE
    )

    if (is.finite(n)) {
      table_data <- utils::head(
        table_data,
        n
      )
    }

    print(
      knitr::kable(
        table_data,
        caption = caption,
        format = "html",
        digits = digits,
        escape = TRUE
      )
    )
  } else {
    cat(
      paste0(
        "\n> Table not found: `",
        path,
        "`. Run the associated script and knit again.\n\n"
      )
    )
  }
}

html_escape <- function(text) {
  text <- gsub(
    "&",
    "&amp;",
    text,
    fixed = TRUE
  )

  text <- gsub(
    "<",
    "&lt;",
    text,
    fixed = TRUE
  )

  text <- gsub(
    ">",
    "&gt;",
    text,
    fixed = TRUE
  )

  text
}

show_script_code <- function(
    path,
    title = NULL
) {
  if (is.null(title)) {
    title <- basename(path)
  }

  if (!file.exists(path)) {
    cat(
      paste0(
        "\n> Source-code file not found: `",
        path,
        "`.\n\n"
      )
    )

    return(
      invisible(NULL)
    )
  }

  code_lines <- readLines(
    path,
    warn = FALSE
  )

  escaped_code <- html_escape(
    paste(
      code_lines,
      collapse = "\n"
    )
  )

  cat(
    paste0(
      "\n<details style='margin:1em 0;'>",
      "<summary><strong>View R code used: ",
      title,
      "</strong></summary>",
      "<pre style='max-height:650px;overflow:auto;",
      "background:#f7f7f7;padding:1em;border:1px solid #ddd;'>",
      "<code class='language-r'>",
      escaped_code,
      "</code></pre>",
      "</details>\n\n"
    )
  )
}

show_file_link <- function(
    path,
    label
) {
  if (file.exists(path)) {
    cat(
      paste0(
        "<p><a href='",
        path,
        "'>",
        label,
        "</a></p>\n"
      )
    )
  }
}
```

# Supplementary figure and table generation

The following five later scripts are lightweight and are run in separate clean
R processes. They create the study-design figure, feature-selection workflow,
HTML-ready model-performance table, shared-candidate-biomarker table and the
final publication-style differential-expression figures.

```{r run-supplementary-scripts}
supplementary_scripts <- c(
  "Scripts/Thesis_Figures/16_Dataset_Study_Design.R",
  "Scripts/Thesis_Figures/17_Feature_Selection_Strategy.R",
  "Scripts/Thesis_Figures/18_Model_Performance_Table.R",
  "Scripts/Thesis_Figures/19_Shared_Candidate_Biomarkers_Table.R",
  "Scripts/Thesis_Figures/20_Publication_Style_DE_Figures.R"
)

dir.create(
  "Results/Logs",
  recursive = TRUE,
  showWarnings = FALSE
)

rscript_command <- file.path(
  R.home("bin"),
  "Rscript"
)

supplementary_status <- lapply(
  supplementary_scripts,
  function(script_path) {
    if (!file.exists(script_path)) {
      return(
        data.frame(
          Script = basename(script_path),
          Status = "Missing",
          Log = NA_character_
        )
      )
    }

    log_path <- file.path(
      "Results/Logs",
      paste0(
        tools::file_path_sans_ext(
          basename(script_path)
        ),
        ".log"
      )
    )

    exit_status <- system2(
      command = rscript_command,
      args = c(
        "--vanilla",
        shQuote(
          normalizePath(
            script_path,
            winslash = "/",
            mustWork = TRUE
          )
        )
      ),
      stdout = log_path,
      stderr = log_path
    )

    data.frame(
      Script = basename(script_path),
      Status = ifelse(
        exit_status == 0,
        "Completed",
        paste0(
          "Failed (exit ",
          exit_status,
          ")"
        )
      ),
      Log = log_path
    )
  }
)

supplementary_status <- dplyr::bind_rows(
  supplementary_status
)

knitr::kable(
  supplementary_status,
  caption = "Supplementary script execution status"
)
```

# Study design and molecular subtypes

**What was done:** The TCGA and CPTAC cohorts, training/test split, analytical
workflow and three harmonised molecular-subtype groups were summarised in one
study-design figure.

```{r study-design-output, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Thesis_Figures/16_Dataset_Study_Design.R",
  "Study design and molecular-subtype figure"
)

show_figure(
  "Figures/Dataset_Slide/Datasets_StudyDesign_MolecularSubtypes_Final.png",
  "Figure 2.1. Datasets, study design and molecular subtypes"
)
```

# Data exploration

**What was done:** The TCGA expression matrix and metadata were loaded, sample
identifiers were checked and the harmonised subtype groups were summarised.

```{r data-overview, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/01_Data_Exploration.R",
  "Data exploration"
)

metadata_path <- paste0(
  "TCGA Data/subtype_tumor_only/",
  "tcga_ucec_subtype_metadata.csv"
)

if (file.exists(metadata_path)) {
  tcga_meta_report <- readr::read_csv(
    metadata_path,
    show_col_types = FALSE
  )

  subtype_summary <- as.data.frame(
    table(
      tcga_meta_report$subtype3
    )
  )

  names(subtype_summary) <- c(
    "Molecular subtype",
    "Number of samples"
  )

  print(
    knitr::kable(
      subtype_summary,
      caption = "TCGA-UCEC three-class subtype distribution",
      format = "html"
    )
  )
}
```

# Principal component analysis

**What was done:** Low-expression and mitochondrial genes were removed,
expression was transformed to log2 CPM, the 1,000 most variable genes were
selected and PCA was used to examine global transcriptional patterns.

```{r pca-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/02_PCA_Analysis.R",
  "Principal component analysis"
)

show_figure(
  "Figures/PCA/PCA_TCGA_Subtypes_Top1000_Combined.png",
  "Principal component analysis of the 1,000 most variable genes"
)

show_table(
  "Figures/PCA/PCA_Variance_Explained.csv",
  "Variance explained by the principal components",
  n = 10
)
```

# Most variable genes

**What was done:** Non-mitochondrial genes were ranked by variance in
normalised expression.

```{r variable-gene-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/03_Top_Variable_Genes.R",
  "Top variable genes"
)

show_figure(
  "Figures/Exploratory/Top20_Most_Variable_Genes.png",
  "Top 20 most variable non-mitochondrial genes"
)

show_table(
  "Results/Exploratory/Top20_Most_Variable_NonMito_Genes.csv",
  "Top variable genes ranked by variance",
  n = 20
)
```

# Exploratory heatmap

**What was done:** The top 20 variable genes were standardised across samples
and visualised with hierarchical clustering and subtype annotations.

```{r exploratory-heatmap, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/04_Heatmap_Top20_Genes.R",
  "Exploratory heatmap"
)

show_figure(
  "Figures/Exploratory/Heatmap_Top20_Variable_Genes.png",
  "Heatmap of the top 20 variable genes"
)
```

# Differential-expression analysis

**What was done:** limma-voom compared CN-high with CN-low, CN-high with
Hypermutated, and CN-low with Hypermutated tumours. Genes were considered
significant at adjusted P < 0.05 and absolute log2 fold change > 1.

```{r de-summary, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/05_Differential_Expression_limma_voom.R",
  "limma-voom differential-expression analysis"
)

show_table(
  "Results/DE_Significant_Gene_Counts.csv",
  "Number of significant genes in each pairwise comparison"
)
```

# Publication-style differential-expression figures

**Why this analysis was included:** Differential-expression counts indicate the
relative scale of transcriptomic divergence between subtype pairs, while
volcano plots show how effect size and statistical significance combine at the
gene level. Together, these outputs help establish which comparisons are most
transcriptionally distinct and provide the biological context for subsequent
enrichment and biomarker analyses.

Script 06 generated the initial diagnostic volcano and count plots. Script 20
uses the same limma-voom result tables to produce the final publication-style
figures, labelled-gene tables and optional MA plots.

```{r publication-de-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/06_Volcano_Plots_limma_voom.R",
  "Initial volcano plots and DEG-count visualisation"
)

show_script_code(
  "Scripts/Thesis_Figures/20_Publication_Style_DE_Figures.R",
  "Final publication-style differential-expression figures"
)

show_figure(
  paste0(
    "Figures/Differential_Expression/",
    "Combined_Volcano_Plots_600dpi.png"
  ),
  paste0(
    "Publication-style volcano plots for the three ",
    "molecular-subtype comparisons"
  )
)

show_figure(
  paste0(
    "Figures/Differential_Expression/",
    "Significant_DEG_Counts_600dpi.png"
  ),
  paste0(
    "Figure 3.3. Number of significantly differentially expressed genes ",
    "across molecular subtype comparisons"
  )
)

show_table(
  paste0(
    "Results/Differential_Expression/",
    "Significant_DEG_Counts_by_Comparison.csv"
  ),
  "Significant DEG counts by comparison and direction"
)

show_table(
  paste0(
    "Results/Differential_Expression/",
    "Total_Significant_DEGs_by_Comparison.csv"
  ),
  "Total significant DEGs in each molecular-subtype comparison"
)

show_table(
  paste0(
    "Results/Differential_Expression/",
    "Volcano_Plot_Labelled_Genes.csv"
  ),
  "Genes labelled in the publication-style volcano plots"
)

if (
  file.exists(
    paste0(
      "Figures/Differential_Expression/",
      "Combined_MA_Plots_600dpi.png"
    )
  )
) {
  show_figure(
    paste0(
      "Figures/Differential_Expression/",
      "Combined_MA_Plots_600dpi.png"
    ),
    "MA plots for the three molecular-subtype comparisons"
  )
}
```

# Top differentially expressed genes

```{r top-de-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/07_Top_DE_Genes.R",
  "Top differentially expressed genes"
)

show_table(
  "Results/Top20_DE_CNhigh_vs_CNlow.csv",
  "Top genes: CN-high versus CN-low",
  n = 10
)

show_table(
  "Results/Top20_DE_CNhigh_vs_Hypermutated.csv",
  "Top genes: CN-high versus Hypermutated",
  n = 10
)

show_table(
  "Results/Top20_DE_CNlow_vs_Hypermutated.csv",
  "Top genes: CN-low versus Hypermutated",
  n = 10
)
```

# Functional enrichment

**What was done:** Significant genes were converted to Entrez identifiers and
used for Gene Ontology Biological Process and KEGG over-representation
analyses.

```{r enrichment-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/08_Functional_Enrichment_GO_KEGG.R",
  "GO and KEGG enrichment"
)

show_figure(
  "Figures/Enrichment/GO_BP_Combined.png",
  "Gene Ontology Biological Process enrichment"
)

show_figure(
  "Figures/Enrichment/KEGG_Combined.png",
  "KEGG pathway enrichment"
)

show_table(
  "Results/Enrichment/Enrichment_Summary.csv",
  "Number of enriched GO terms and KEGG pathways"
)
```

# Machine-learning data preparation

**What was done:** TCGA samples were divided into 186 training samples and 46
held-out test samples. Filtering, selection of 1,000 genes and standardisation
were learned using the training cohort only.

```{r ml-preparation, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/09_ML_Data_Preparation.R",
  "Machine-learning data preparation"
)

show_table(
  "Results/ML/ML_Preprocessing_Summary.csv",
  "Machine-learning preprocessing summary"
)
```

# Feature-selection strategy

**What was done:** The same 1,000 training-selected predictors were supplied
independently to multinomial LASSO and Random Forest. LASSO retained 56 genes
at lambda.min, Random Forest contributed its top 30 permutation-importance
genes and their intersection produced 11 shared candidate biomarkers.

```{r feature-selection-output, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Thesis_Figures/17_Feature_Selection_Strategy.R",
  "Feature-selection strategy"
)

show_figure(
  "Figures/Feature_Selection/Feature_Selection_Workflow_Final.png",
  "Figure 2.2. Feature-selection strategy used for candidate biomarker prioritisation"
)
```

# Multinomial LASSO

```{r lasso-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/10_LASSO_Classification.R",
  "Multinomial LASSO training and evaluation"
)

show_figure(
  "Figures/ML/LASSO_CrossValidation.png",
  "Five-fold cross-validation for multinomial LASSO"
)

show_figure(
  "Figures/ML/LASSO_ConfusionMatrix_lambda_min.png",
  "LASSO held-out test confusion matrix at lambda.min"
)

show_figure(
  "Figures/ML/LASSO_Top_Selected_Genes.png",
  "Top genes retained by the LASSO model"
)

show_table(
  "Results/ML/LASSO_Overall_Metrics.csv",
  "LASSO overall held-out test metrics"
)
```

# Random Forest

```{r random-forest-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/11_Random_Forest_Classification.R",
  "Random Forest training and permutation importance"
)

show_figure(
  "Figures/ML/RandomForest_CrossValidation.png",
  "Random Forest cross-validation performance"
)

show_figure(
  "Figures/ML/RandomForest_ConfusionMatrix.png",
  "Random Forest held-out test confusion matrix"
)

show_figure(
  "Figures/ML/RandomForest_Top_Genes.png",
  "Figure 3.7. Random Forest feature importance for subtype classification"
)

show_table(
  "Results/ML/RandomForest_Overall_Metrics.csv",
  "Random Forest overall held-out test metrics"
)
```

# Internal model comparison

```{r internal-comparison, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/12_Model_Comparison.R",
  "Internal LASSO and Random Forest comparison"
)

show_figure(
  "Figures/ML/LASSO_vs_RandomForest_Performance.png",
  "Held-out TCGA performance: LASSO versus Random Forest"
)

show_table(
  "Results/ML/Model_Comparison_Primary.csv",
  "Primary internal model comparison"
)
```

# Independent CPTAC evaluation

**What was done:** The TCGA-trained models were applied to CPTAC-UCEC without
retraining. This is an external transfer evaluation rather than clinical
validation.

```{r external-evaluation, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/13_CPTAC_External_Evaluation.R",
  "Independent CPTAC external evaluation"
)

show_figure(
  "Figures/External_Validation/Internal_vs_External_Accuracy.png",
  "Internal TCGA and external CPTAC accuracy"
)

show_figure(
  "Figures/External_Validation/CPTAC_LASSO_ConfusionMatrix.png",
  "CPTAC LASSO confusion matrix"
)

show_figure(
  "Figures/External_Validation/CPTAC_RandomForest_ConfusionMatrix.png",
  "CPTAC Random Forest confusion matrix"
)

show_table(
  "Results/External_Validation/CPTAC_Model_Metrics.csv",
  "CPTAC external evaluation metrics"
)

show_table(
  "Results/External_Validation/CPTAC_Gene_and_Scale_Compatibility.csv",
  "CPTAC gene-coverage and scale compatibility"
)
```

# Internal and external model-performance table

The same data used to create the editable Word table are written to CSV and
displayed directly below, so the table is visible inside the HTML.

```{r performance-table-output, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Thesis_Figures/18_Model_Performance_Table.R",
  "Internal and external model-performance table"
)

show_table(
  "Results/Tables/Model_Performance_Table.csv",
  paste0(
    "Internal and external performance of LASSO and Random Forest models ",
    "for UCEC molecular subtype classification"
  ),
  digits = 3
)

show_file_link(
  "Results/Tables/Model_Performance_Table_Editable.docx",
  "Open the editable Word version of this table"
)
```

# Candidate-biomarker prioritisation

**What was done:** Genes retained by LASSO were compared with the top 30 Random
Forest genes. Differential-expression and biological evidence were added to
prioritise candidate biomarkers.

```{r biomarker-results, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/14_Biological_Interpretation.R",
  "Biological interpretation and candidate-biomarker prioritisation"
)

show_script_code(
  "Scripts/Thesis_Figures/19_Shared_Candidate_Biomarkers_Table.R",
  "Shared candidate-biomarker table"
)

show_figure(
  "Figures/Biological_Interpretation/Priority_Biomarker_Heatmap_600dpi.png",
  "Expression heatmap for prioritised candidate biomarkers"
)

show_figure(
  "Figures/Biological_Interpretation/Priority_Gene_Expression_Boxplots.png",
  "Priority candidate-gene expression across TCGA-UCEC molecular subtypes"
)

show_figure(
  "Figures/Biological_Interpretation/Priority_Biomarker_Evidence_Scores.png",
  "Evidence scores for prioritised candidate biomarkers"
)

show_table(
  "Results/ML/Shared_Candidate_Biomarkers_Detailed.csv",
  "Eleven shared candidate biomarkers supported by LASSO and Random Forest",
  digits = 4
)

show_table(
  "Results/Biological_Interpretation/ML_Gene_Overlap_Summary.csv",
  "Overlap between LASSO and Random Forest gene selections"
)

show_table(
  "Results/Biological_Interpretation/Top10_Biomarkers_for_Presentation.csv",
  "Top candidate biomarkers for presentation",
  n = 10
)
```

# Final integrated summary

```{r final-summary, results='asis', echo=FALSE}
show_script_code(
  "Scripts/Final/15_Final_Figures_and_Summary_Tables.R",
  "Final figures and summary tables"
)

show_figure(
  "Figures/Final/Combined_ML_and_Biomarker_Summary.png",
  "Combined machine-learning and biomarker summary"
)

show_figure(
  "Figures/Final/Internal_vs_External_Model_Accuracy.png",
  "Final internal and external model-accuracy comparison"
)

show_table(
  "Results/Final/Final_Project_Summary.csv",
  "Final project summary",
  n = 20
)

show_table(
  "Results/Final/Key_Findings_for_Presentation.csv",
  "Key findings prepared for presentation",
  n = 20
)
```

# Output audit

```{r report-output-audit}
report_outputs <- c(
  "Figures/Dataset_Slide/Datasets_StudyDesign_MolecularSubtypes_Final.png",
  "Figures/PCA/PCA_TCGA_Subtypes_Top1000_Combined.png",
  "Figures/Exploratory/Top20_Most_Variable_Genes.png",
  "Figures/Exploratory/Heatmap_Top20_Variable_Genes.png",
  "Figures/Differential_Expression/Combined_Volcano_Plots_600dpi.png",
  "Figures/Differential_Expression/Significant_DEG_Counts_600dpi.png",
  "Results/Differential_Expression/Significant_DEG_Counts_by_Comparison.csv",
  "Results/Differential_Expression/Total_Significant_DEGs_by_Comparison.csv",
  "Figures/Enrichment/GO_BP_Combined.png",
  "Figures/Enrichment/KEGG_Combined.png",
  "Figures/Feature_Selection/Feature_Selection_Workflow_Final.png",
  "Figures/ML/LASSO_CrossValidation.png",
  "Figures/ML/RandomForest_Top_Genes.png",
  "Figures/ML/LASSO_vs_RandomForest_Performance.png",
  "Figures/External_Validation/Internal_vs_External_Accuracy.png",
  "Figures/Biological_Interpretation/Priority_Biomarker_Heatmap_600dpi.png",
  "Figures/Biological_Interpretation/Priority_Gene_Expression_Boxplots.png",
  "Figures/Final/Combined_ML_and_Biomarker_Summary.png",
  "Results/Tables/Model_Performance_Table.csv",
  "Results/ML/Shared_Candidate_Biomarkers_Detailed.csv"
)

output_audit <- data.frame(
  Output = report_outputs,
  Exists = file.exists(report_outputs),
  Size_KB = ifelse(
    file.exists(report_outputs),
    round(
      file.info(report_outputs)$size / 1024,
      1
    ),
    NA_real_
  )
)

knitr::kable(
  output_audit,
  caption = "Final figure and table output audit",
  format = "html"
)
```

# Reproducibility information

The heavy workflow remains in the 15 numbered scripts under `Scripts/Final`.
The five later scripts are stored under `Scripts/Thesis_Figures`. All source
code is shown inside collapsible sections in this HTML, including the
publication-style differential-expression script that generates Figure 3.3.

```{r session-information}
sessionInfo()
```
