---
title: "Reproducible Analysis: UCEC Molecular Subtype Classification (Low-Memory Run)"
author: "Student ID: 3173204"
date: "31 July 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
---

# Purpose

This document runs the complete UCEC transcriptomic analysis in a fixed and
reproducible order. It covers data checks, exploratory analysis, differential
expression, functional enrichment, machine-learning development, independent
CPTAC evaluation, biomarker prioritisation and final summary outputs.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 150
)

options(stringsAsFactors = FALSE)
set.seed(123)

project_root <- normalizePath(
  ".",
  winslash = "/",
  mustWork = TRUE
)

knitr::opts_knit$set(
  root.dir = project_root
)

cat("Project root:", project_root, "\n")
cat("R version:", R.version.string, "\n")
```

# 1. Project setup

```{r project-setup}
folders_to_create <- c(
  "Scripts/Final",
  "Figures",
  "Results",
  "Tables"
)

invisible(
  lapply(
    folders_to_create,
    dir.create,
    recursive = TRUE,
    showWarnings = FALSE
  )
)

required_data_files <- c(
  paste0(
    "TCGA Data/subtype_tumor_only/",
    "tcga_ucec_subtype_counts_geneSymbol.csv"
  ),
  paste0(
    "TCGA Data/subtype_tumor_only/",
    "tcga_ucec_subtype_metadata.csv"
  ),
  paste0(
    "CPTAC Data/subtype_tumor_only/",
    "cptac_ucec_subtype_rna.csv"
  ),
  paste0(
    "CPTAC Data/subtype_tumor_only/",
    "cptac_ucec_subtype_metadata.csv"
  )
)

data_status <- data.frame(
  file = required_data_files,
  exists = file.exists(required_data_files)
)

knitr::kable(
  data_status,
  caption = "Required data-file check"
)

if (!all(data_status$exists)) {
  stop(
    "One or more required data files are missing. ",
    "Place the TCGA and CPTAC files in the paths shown above."
  )
}
```

# 2. Package check

```{r package-check}
required_packages <- c(
  "tidyverse",
  "readr",
  "edgeR",
  "limma",
  "caret",
  "glmnet",
  "Matrix",
  "ranger",
  "clusterProfiler",
  "AnnotationDbi",
  "org.Hs.eg.db",
  "enrichplot",
  "pheatmap",
  "patchwork",
  "grid"
)

package_status <- data.frame(
  package = required_packages,
  installed = vapply(
    required_packages,
    requireNamespace,
    quietly = TRUE,
    FUN.VALUE = logical(1)
  )
)

knitr::kable(
  package_status,
  caption = "Required R-package check"
)

if (!all(package_status$installed)) {
  stop(
    "Install the missing packages before knitting: ",
    paste(
      package_status$package[
        !package_status$installed
      ],
      collapse = ", "
    )
  )
}
```

# 3. Script inventory

Scripts are executed alphabetically; the two-digit prefixes therefore determine
the analytical order.

```{r script-inventory}
final_scripts <- sort(
  list.files(
    path = "Scripts/Final",
    pattern = "^[0-9]{2}_.+\\.R$",
    full.names = TRUE
  )
)

if (length(final_scripts) != 15) {
  stop(
    "Expected 15 numbered scripts in Scripts/Final, but found ",
    length(final_scripts),
    "."
  )
}

script_inventory <- data.frame(
  order = seq_along(final_scripts),
  script = basename(final_scripts),
  size_kb = round(
    file.info(final_scripts)$size / 1024,
    1
  )
)

knitr::kable(
  script_inventory,
  caption = "Scripts included in the workflow"
)
```

# 4. Run the complete analysis

To reduce memory use, every numbered script is run in a separate clean R
process. This prevents large expression matrices and fitted models from
accumulating in the knitting session. A separate log is saved for every script.

```{r run-analysis, echo=TRUE, results='asis'}
dir.create(
  "Results/Logs",
  recursive = TRUE,
  showWarnings = FALSE
)

rscript_command <- file.path(
  R.home("bin"),
  "Rscript"
)

if (!file.exists(rscript_command)) {
  stop(
    "Rscript executable was not found at: ",
    rscript_command
  )
}

run_script_separately <- function(
    path,
    script_number
) {
  script_name <- basename(path)

  log_stub <- tools::file_path_sans_ext(
    script_name
  )

  log_file <- file.path(
    "Results/Logs",
    paste0(
      sprintf("%02d", script_number),
      "_",
      log_stub,
      ".log"
    )
  )

  cat(
    paste0(
      "\n\n**Running ",
      script_name,
      "...**\n\n"
    )
  )

  # Each child process starts with --vanilla, so it does not restore .RData.
  exit_status <- system2(
    command = rscript_command,
    args = c(
      "--vanilla",
      shQuote(
        normalizePath(
          path,
          winslash = "/",
          mustWork = TRUE
        )
      )
    ),
    stdout = log_file,
    stderr = log_file
  )

  if (!identical(exit_status, 0L)) {
    log_lines <- readLines(
      log_file,
      warn = FALSE
    )

    log_tail <- tail(
      log_lines,
      60
    )

    stop(
      paste0(
        "Script failed: ",
        script_name,
        "\n\nLast lines from ",
        log_file,
        ":\n",
        paste(
          log_tail,
          collapse = "\n"
        )
      ),
      call. = FALSE
    )
  }

  cat(
    paste0(
      "Completed successfully. Log: `",
      log_file,
      "`\n\n"
    )
  )

  invisible(gc())
}

for (script_number in seq_along(final_scripts)) {
  run_script_separately(
    path = final_scripts[[script_number]],
    script_number = script_number
  )
}
```

# 5. Core reproducibility checks

```{r key-result-checks}
x_train_check <- readRDS(
  "Results/ML/x_train_scaled.rds"
)

x_test_check <- readRDS(
  "Results/ML/x_test_scaled.rds"
)

y_train_check <- readRDS(
  "Results/ML/y_train.rds"
)

y_test_check <- readRDS(
  "Results/ML/y_test.rds"
)

shared_gene_check <- readr::read_csv(
  "Results/ML/LASSO_RandomForest_Shared_Genes.csv",
  show_col_types = FALSE
)

cptac_metric_check <- readr::read_csv(
  "Results/External_Validation/CPTAC_Model_Metrics.csv",
  show_col_types = FALSE
)

stopifnot(
  nrow(x_train_check) == 186,
  nrow(x_test_check) == 46,
  length(y_train_check) == 186,
  length(y_test_check) == 46,
  ncol(x_train_check) == 1000,
  ncol(x_test_check) == 1000,
  nrow(shared_gene_check) == 11,
  nrow(cptac_metric_check) >= 2
)

key_checks <- data.frame(
  Check = c(
    "Training samples",
    "Held-out test samples",
    "Training-selected predictors",
    "Shared LASSO/RF candidate genes",
    "CPTAC model rows"
  ),
  Observed = c(
    nrow(x_train_check),
    nrow(x_test_check),
    ncol(x_train_check),
    nrow(shared_gene_check),
    nrow(cptac_metric_check)
  ),
  Expected = c(
    186,
    46,
    1000,
    11,
    2
  )
)

knitr::kable(
  key_checks,
  caption = "Core result checks"
)
```

# 6. Final output audit

```{r output-audit}
expected_outputs <- c(
  "Figures/PCA/PCA_TCGA_Subtypes_Top1000_Combined.png",
  "Figures/DE/Volcano_Plots_Combined.png",
  "Figures/DE/DEG_Counts_by_Comparison.png",
  "Figures/Enrichment/GO_BP_Combined.png",
  "Figures/Enrichment/KEGG_Combined.png",
  "Figures/ML/LASSO_CrossValidation.png",
  "Figures/ML/RandomForest_Top_Genes.png",
  "Results/ML/Model_Comparison_Primary.csv",
  "Results/External_Validation/CPTAC_Model_Metrics.csv",
  paste0(
    "Results/Biological_Interpretation/",
    "All_ML_DE_Biomarker_Evidence.csv"
  ),
  paste0(
    "Figures/Biological_Interpretation/",
    "Priority_Biomarker_Heatmap_600dpi.png"
  )
)

output_status <- data.frame(
  output = expected_outputs,
  exists = file.exists(expected_outputs),
  size_kb = ifelse(
    file.exists(expected_outputs),
    round(
      file.info(expected_outputs)$size / 1024,
      1
    ),
    NA_real_
  )
)

knitr::kable(
  output_status,
  caption = "Final output audit"
)

if (!all(output_status$exists)) {
  warning(
    "Some expected outputs were not regenerated. ",
    "Review the output-audit table."
  )
}
```

# 7. Data availability

The TCGA-UCEC and CPTAC-UCEC datasets may be subject to access or
redistribution conditions. The code, expected directory structure and input
filenames are supplied so that authorised users can reproduce the analysis
after obtaining the datasets from the appropriate sources.

# 8. Computational environment

```{r session-information}
sessionInfo()
```
